Exploratory Data Analysis of Loan Data from Prosper
_Barbara Stempien_
========================================================

Prosper Marketplace is America's first peer-to-peer lending marketplace, with over \$7 billion in funded loans. Borrowers request personal loans on Prosper and investors (individual or institutional) can fund anywhere from \$2,000 to \$35,000 per loan request. Investors can consider borrowers’ credit scores, ratings, and histories and the category of the loan. Prosper handles the servicing of the loan and collects and distributes borrower payments and interest back to the loan investors.

Prosper verifies borrowers' identities and select personal data before funding loans and manages all stages of loan servicing. Prosper's unsecured personal loans are fully amortized over a period of three or five years, with no pre-payment penalties. Prosper generates revenue by collecting a one-time fee on funded loans from borrowers and assessing an annual loan servicing fee to investors.

Prosper publishes performance statistics on its website and all market data is available to the public for analysis. Prosper loan data set contains 113,937 loans with 81 variables on each loan, 
including loan amount, borrower rate (or interest rate), current loan status, 
borrower income, borrower employment status, borrower credit history, and 
the latest payment information.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='plots/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r packages}
# Load packages
library(ggplot2)
library(repr)
library(ggthemes)
library(dplyr)
library(maps)
library(mapdata)
library(openintro)
library(waffle)
library(ggcorrplot)
library(scales)
library(ggpubr)
library(gridExtra)
```

```{r Load_the_Data}
# load the prosper loan dataset
loan_dataset <- read.csv('data/prosperLoanData.csv', na.strings = c("", "NA"))
```

```{r, Data Wrangling}

# subset the dataframe to the following columns
subset <- c("LoanStatus",
            "ListingCreationDate",
            "ClosedDate",
            "ListingCategory..numeric.",
            "Term",
            "BorrowerRate",
            "LoanOriginalAmount",
            "MonthlyLoanPayment",
            "BorrowerState",
            "IsBorrowerHomeowner",  
            "Occupation",
            "EmploymentStatus",
            "StatedMonthlyIncome",
            "DebtToIncomeRatio",
            "ProsperRating..Alpha.",
            "ProsperScore",           
            "OpenCreditLines",
            "TotalCreditLinespast7years",
            "OpenRevolvingAccounts",
            "OpenRevolvingMonthlyPayment",
            "CurrentDelinquencies",
            "AmountDelinquent",
            "DelinquenciesLast7Years",
            "PublicRecordsLast10Years",
            "RevolvingCreditBalance",
            "BankcardUtilization",
            "LenderYield",
            "Investors"
           )

# subset dataframe
df <- loan_dataset[subset]

# replace NAs with 0
df[is.na(df)] <- 0
```
``` {r ,Clean and Tidy dataset}

# convert Loan Status to factor with 7 levels
# https://stackoverflow.com/a/19410249/7382214
levels(df$LoanStatus) <- list(Current = "Current",
                              Completed = "Completed",
                              "Charged Off" = "Chargedoff",
                              Defaulted = "Defaulted",
                              "Past Due" = c("Past Due (>120 days)",
                                        "Past Due (1-15 days)",
                                        "Past Due (16-30 days)",
                                        "Past Due (31-60 days)",
                                        "Past Due (61-90 days)",
                                        "Past Due (91-120 days)"),
                              "Final Payment" = "FinalPaymentInProgress",
                              Cancelled = "Cancelled")

# convert Listing Creation Date variable to date format
df$ListingCreationDate <- as.Date(df$ListingCreationDate)

# convert Term to ordered factor with 3 levels
df$Term <- factor(df$Term, levels=c(12,36,60), ordered=TRUE)

# convert Closed Date variable to date format
df$ClosedDate <- as.Date(df$ClosedDate)

# convert Prosper Rating to ordered factor with 8 levels
df$ProsperRating..Alpha. <- factor(df$ProsperRating..Alpha., levels=c("HR", "E", "D", "C","B","A","AA"), ordered=TRUE)

# convert Prosper Score to ordered factor with 12 levels
df$ProsperScore <- factor(df$ProsperScore, levels=c(0,1,2,3,4,5,6,7,8,9,10,11), ordered=TRUE)

# add column with Listing Category labels based on the documentation
df$ListingCategory <- factor(df$ListingCategory..numeric., labels=c('Not Available',
                                                                    'Debt Consolidation',
                                                                    'Home Improvement',
                                                                    'Business',
                                                                    'Personal Loan',
                                                                    'Student Use',
                                                                    'Auto',
                                                                    'Other',
                                                                    'Baby&Adoption',
                                                                    'Boat',
                                                                    'Cosmetic Procedure',
                                                                    'Engagement Ring',
                                                                    'Green Loans',
                                                                    'Household Expenses',
                                                                    'Large Purchases',
                                                                    'Medical/Dental',
                                                                    'Motorcycle',
                                                                    'RV',
                                                                    'Taxes',
                                                                    'Vacation',
                                                                    'Wedding Loans'))

# move ListingCategory column next to the ListingCategory..numeric.
df <- df[c(1:4,29,5:28)]

# convert State abbreviation to full name
df$BorrowerState <- abbr2state(df$BorrowerState)

# convert Is Borrower Homeowner to logical type
df$IsBorrowerHomeowner <- as.logical(df$IsBorrowerHomeowner)

# replace NA in Occupation with Not Available
levels(df$Occupation)<-c(levels(df$Occupation),"Not Available") 
df$Occupation[is.na(df$Occupation)] <- "Not Available"

# replace NA in Employment Status with Not Available
df$EmploymentStatus[is.na(df$EmploymentStatus)] <- "Not available"

```

# Univariate Plots Section

> We will begin the analysis by looking at the data contained in the data set.

```{r echo=FALSE, View dataset}
head(loan_dataset)
```

```{r echo=FALSE, Summary}
summary(df)
```

First, I want to see when and how many listings were created.

```{r}
# Group Listings by year and month
creation_date <- df %>% count(df$ListingCreationDate) %>%
                setNames(.,c("date", "count")) %>%
                mutate(month = as.Date(cut(date, breaks = "month")))  %>%
                group_by(month) %>% 
                summarise(total = sum(count)) 

# Plot Listing Creation Date
ggplot(creation_date, aes(x=month, y=total)) +
      geom_line(color = "#3F5D7D", size = 0.75) + 
      geom_point(color = "#3F5D7D", size = 3) +
      scale_x_date(date_labels = "%b-%Y", date_breaks="4 months") +
      labs(title = "When was the listing created?") + 
      scale_y_continuous(limits = c(0,6000), breaks = seq(0, 6000, 1000)) +
      theme(plot.title = element_text(size = 18),
            axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.y = element_text(size=10),
            axis.text.x = element_text(angle = 45, hjust = 1, size=10),
            panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5))

```

From the graph above, we can see that the number of loans in Prosper has increased over time. In 2008 there was a collapse in the number of listings, however, since 2009, the number has been constantly growing.

```{r}
# Loan Original Amount statistics
summary(df$LoanOriginalAmount)

# plot Loan Original Amount
ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(binwidth = 1000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1000, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(breaks = seq(0, 36000, 1000), labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,17500), breaks = seq(0, 17500, 2500), expand = c(0,0)) +
  labs(title="How high is the original loan amount (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(LoanOriginalAmount)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 10000, y = 17000, 
           label = paste("Mean: $", round(mean(df$LoanOriginalAmount))), size = 4, color = "#ff7f0e")
```
 
Another interesting feature is the amount of loans. In the above graph we can see that most loans are granted for amounts from \$4,000 to \$12,000. The average loan amount is \$8337.


```{r}
# Term
term <- df %>% count(df$Term) %>%
                    setNames(.,c("term", "count"))

# plot term
waffle(c("12 months" = term$count[term$term == 12] / 500, 
         "36 months" = term$count[term$term == 36] / 500,
         "60 months" = term$count[term$term == 60] / 500),
       rows = 10, 
       colors = c("#d3d3d3", "#a9bacc", "#3F5D7D"), 
       size = 0.5, 
       pad = 1, 
       legend_pos = "bottom", 
       title = "How long is a loan term?",
       xlab = "1 square = 500 customers")
```

The vast majority of loans is granted for 36 months. Only a fraction of all loans is granted for 12 months.


```{r}

# Listing Category
listing_category <- df %>% count(df$ListingCategory) %>%
                    setNames(.,c("category", "count"))

# plot listing category
ggplot(listing_category, aes(x = category, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  labs(title = "What is the listing's category?") + 
  scale_y_continuous(limits = c(0,65000), breaks = seq(0, 65000, 5000), expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5))
```

Consolidation of debts is the most common reason for a loan. We do not have information about the category for a large number of loans - 16,000. The second most common category is Other, followed by Home Improvement and Business.

```{r}

# Loan Status 
loan_status <- df %>% count(df$LoanStatus) %>%
                setNames(.,c("status", "count")) %>%
                arrange(desc(count))

# plot loan status
ggplot(loan_status, aes(x = status, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.25) +
  labs(title = "What is the loan status?") + 
  scale_y_continuous(limits = c(0,60000), breaks = seq(0, 55000, 5000), expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  rotate()

```

Most of the loans in the dataset are currently active (56,000). Over 38,000, was paid off. Around 16 thousand were charged off or defaulted.

```{r}
# Borrower Rate
summary(df$BorrowerRate)
        
borrower_rate <- df %>% count(df$BorrowerRate) %>%
                    set_colnames(c("rate", "count"))

ggplot(aes(x = BorrowerRate * 100), data = df) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(breaks = seq(0, 50, 1), labels=dollar_format(prefix="", suffix="%")) +
  scale_y_continuous(limits = c(0,6500), breaks = seq(0, 6500, 1000), expand = c(0,0)) +
  labs(title="What is the borrower's rate?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(BorrowerRate) * 100),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 22, y = 6250, 
           label = paste("Mean:", round(mean(df$BorrowerRate * 100),2), "%"), size = 4, color = "#ff7f0e")
```

As for the borrower's rate, it is 0.1928 on average. Although a considerable amount of loans is granted at the borrower's rate of 0.14, 0.15, 0.18, a lot of loans are granted at a much higher rate, for instance 0.32 (5914 loans).

```{r}
# Monthly Loan Payment statistics
summary(df$MonthlyLoanPayment)

# plot Monthly Loan Payment 
ggplot(aes(x = MonthlyLoanPayment), data = df) +
  geom_histogram(binwidth = 100, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 100, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(breaks = seq(0, 2300, 100), labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,30000), breaks = seq(0, 30000, 5000), expand = c(0,0)) +
  labs(title="How high is the monthly loan payment (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(MonthlyLoanPayment)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 400, y = 29000, 
           label = paste("Mean: $", round(mean(df$MonthlyLoanPayment))), size = 4, color = "#ff7f0e")
```

The average loan installment is \$272. Most loan installments are in the \$130 - \$370 range.

```{r}
# group loans by borrower's state
loans <- df %>% count(df$BorrowerState) %>%
                setNames(.,c("region", "count"))
loans$region <- tolower(loans$region)

# get polygon data for states
states <- map_data("state")

# merge datasets
loans_states <- inner_join(states, loans, by = "region")

# plot borrowers by state
ggplot(data = loans_states, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "#000000", fill = "#d3d3d3") + 
  geom_polygon(data = loans_states, aes(fill = count), color = "#ffffff") +
  geom_polygon(color = "#000000", fill = NA) +
  labs(title = "Where does the borrower live?") + 
  theme_bw() +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank()) + 
  scale_fill_gradient(trans = "log10", low = "#d3d3d3", high = "#3F5D7D")
```

The largest number of loans is granted to borrowers living in the California.

```{r}
# IsBorrowerHomeowner variable
homeowner <- df %>% count(df$IsBorrowerHomeowner) %>%
                setNames(.,c("owner", "count"))

# plot home ownership
waffle(c(No = homeowner$count[homeowner$owner == "FALSE"] / 1000, 
         Yes = homeowner$count[homeowner$owner == "TRUE"] / 1000), 
       rows = 6, 
       colors = c("#d3d3d3","#3F5D7D"), 
       size = 0.5, pad = 1, 
       legend_pos = "bottom", 
       title = "Is Borrower a Homeowner?",
       xlab = "1 square = 1000 customers")
```

The share of homeowners in the total number of borrowers is around 50%.

```{r}
# Employment Status
employment <- df %>% count(df$EmploymentStatus) %>%
                    setNames(.,c("employment", "count")) %>%
                    arrange(desc(count))

# plot Employment Status
ggplot(employment, aes(x = employment, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.5) +
  labs(title = "What is the borrower's employment status?") + 
  scale_y_continuous(limits = c(0,75000), breaks = seq(0, 70000, 5000), expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) +
  rotate()
```

The vast majority of borrowers are currently employed. Only a fraction of borrowers do not have a job or work part-time.

```{r}
# Occupation
occupation <- df %>% count(df$Occupation) %>%
                    setNames(.,c("occupation", "count")) %>%
                    arrange(desc(count))

# plot occupation
ggplot(occupation[1:34,], aes(x = occupation, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  labs(title = "What is the borrower's occupation?") + 
  scale_y_continuous(limits = c(0,35000), breaks = seq(0, 35000, 5000), expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5))

ggplot(occupation[35:67,], aes(x = occupation, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  scale_y_continuous(limits = c(0,35000), breaks = seq(0, 35000, 5000), expand = c(0,0)) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5))
```

Unfortunately, for the majority of loans, there is no information about the borrower's profession. A significant part of borrowers are professionals, programmers or executives.

```{r}
# Stated Monthly Income
summary(df$StatedMonthlyIncome)

# plot Stated Monthly Income
ggplot(aes(x = StatedMonthlyIncome), data = df) +
  geom_histogram(binwidth = 1000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1000, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(-999, quantile(df$StatedMonthlyIncome, 0.99)), 
                     breaks = seq(0, 20000, 1000),
                     expand = c(0,0),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,20000), 
                     breaks = seq(0, 20000, 2500), 
                     expand = c(0,0)) +
  labs(title="What income was delcared by the borrower (monthly, USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(StatedMonthlyIncome)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 6500, y = 19500, 
           label = paste("Mean: $", round(mean(df$StatedMonthlyIncome))), size = 4, color = "#ff7f0e")
```

The average monthly income of a borrower is \$5608. The vast majority of borrowers earn in the period between \$3200 - \$6825.

```{r}
# Debt To Income Ratio
summary(df$DebtToIncomeRatio)

# plot debt to income ratio
ggplot(aes(x = DebtToIncomeRatio), data = df) +
  geom_histogram(binwidth = 0.05, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 0.05, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(-0.05, quantile(df$DebtToIncomeRatio, 0.99)), breaks = seq(0, 2, 0.05), expand = c(0,0)) +
  scale_y_continuous(limits = c(0,20000), breaks = seq(0, 20000, 2500), expand = c(0,0)) +
  labs(title="What is the borrower's debt to income ratio?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(DebtToIncomeRatio)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 0.29, y = 19000, 
           label = paste("Mean:", round(mean(df$DebtToIncomeRatio),2)), size = 4, color = "#ff7f0e")
```

It is interesting to compare income to debt. As we can see in the plot above, the average debt to income ratio is 0.2552 and the vast majority of borrowers are in the range of 0.1300 - 0.3100.

```{r}
# Delinquencies
limit <- quantile(df$DelinquenciesLast7Years, 0.99)

ggplot(data = df, aes(x = DelinquenciesLast7Years)) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(-1, limit), breaks = seq(0, limit, 1)) +
  scale_y_continuous(limits = c(0,80000), breaks = seq(0, 80000, 5000), expand = c(0,0)) +
  labs(title = "How many delinquencies did borrower have in past 7 years??") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(DelinquenciesLast7Years)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 6, y = 75000, 
           label = paste("Mean:", round(mean(df$DelinquenciesLast7Years))), size = 4, color = "#ff7f0e")
```

Another important aspect when assessing credit risk is the number of delinquencies that the borrower had in recent years. As we can see in the plot above, most borrowers did not have any delinquencies in the last 7 years.

```{r}
# Public Records
limit <- quantile(df$PublicRecordsLast10Years, 0.99)

ggplot(data = df, aes(x = PublicRecordsLast10Years)) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(-1, limit), breaks = seq(0, limit, 1)) +
  scale_y_continuous(limits = c(0,90000), breaks = seq(0, 90000, 5000), expand = c(0,0)) +
  labs(title = "How many public records did borrower have in past 10 years??") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(PublicRecordsLast10Years)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 6, y = 75000, 
           label = paste("Mean:", round(mean(df$PublicRecordsLast10Years))), size = 4, color = "#ff7f0e")
```

We see similar situations when it comes to public records from the last 10 years. For a significant number of borrowers, it is 0.

```{r}
# Bank card Utilization
summary(df$BankcardUtilization)

ggplot(aes(x = BankcardUtilization * 100), data = df) +
  geom_histogram(binwidth = 5, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 5, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(-3,125), breaks = seq(0, 125, 5), expand = c(0,0), labels=dollar_format(prefix="", suffix="%")) +
  scale_y_continuous(limits = c(0,18000), breaks = seq(0, 18000, 2500), expand = c(0,0)) +
  labs(title="What is the borrower's bankcard utilization?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(BankcardUtilization * 100)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 58, y = 16000, 
           label = paste("Mean:", round(mean(df$BankcardUtilization * 100),2), "%"), size = 4, color = "#ff7f0e")
```

Utilization of a credit card is also an important factor in risk assessment. As we can see in the plot above, the average credit card utilization is 52%, and most borrowers used between 23% and 82% of the available resources on the credit card.

```{r}
# Revolving Credit Balance
summary(df$RevolvingCreditBalance)

# Revolving Credit Balance
ggplot(aes(x = RevolvingCreditBalance), data = df) +
  geom_histogram(binwidth = 5000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 5000, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits = c(0, quantile(df$RevolvingCreditBalance, 0.99)), 
                     breaks = seq(0, 150000, 5000),
                     expand = c(0,0),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,30000), 
                     breaks = seq(0, 30000, 2500), 
                     expand = c(0,0)) +
  labs(title="What is the amount of borrower's revolving credit balance (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(RevolvingCreditBalance)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 24000, y = 28000, 
           label = paste("Mean: $", round(mean(df$RevolvingCreditBalance))), size = 4, color = "#ff7f0e")
```

Another value that can help us analyze the financial status of the borrower is revelving credit balance. As we can see, the average value is \$16,424, and most borrowers are in the \$2091 - \$18254 range.


```{r}
# Prosper Rating
prosper_rating <- df %>% count(df$ProsperRating..Alpha.) %>%
                    setNames(.,c("rating", "count")) %>%
                    arrange(desc(count))

# Prosper Rating
ggplot(prosper_rating, aes(x = rating, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.5) +
  labs(title = "What is the borrower's prosper rating?") + 
  scale_y_continuous(limits = c(0,31000), breaks = seq(0, 31000, 2500), expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) +
  rotate()
```

Based on these data, and several other factors, Prosper assessed the credit risk associated with each borrower. As we can see on the graph above, the most common risk is C, then B. For a large number of borrowers, we do not have any information about the risk, as this information was not collected before 2009.

```{r}
# Number of investors
summary(df$Investors)

ggplot(aes(x = Investors), data = df) +
  geom_density(color = "#ffffff", fill = "#3F5D7D") + 
  scale_x_continuous(limits = c(0,650), breaks = seq(0, 650, 25)) +
  labs(title="What is the number of investors?") + 
  xlab('Number of investors') +
  ylab('Density') +
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(Investors)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 110, y = 0.015, 
           label = paste("Mean:", round(mean(df$Investors),0)), size = 4, color = "#ff7f0e")
```

The average number of investors per loan is 80. Most of the loans have from 1 to 115 investors. The highest number of investors per loan was 1189.

```{r}
# Lender Yield
summary(df$LenderYield)

ggplot(aes(x = LenderYield), data = df) +
  geom_histogram(binwidth = 0.025, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 0.025, aes(y=..count.., label=..count..,), geom="text", vjust=-.5, size=3) +
  scale_x_continuous(limits=c(0,0.37),breaks = seq(0, 0.35, 0.025)) +
  scale_y_continuous(breaks = seq(0, 14000, 2000)) +
  labs(title="What is the lender yield?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", linetype = "dashed", size=0.5)) + 
  geom_vline(aes(xintercept = mean(LenderYield)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 0.205, y = 14000, 
           label = paste("Mean:", round(mean(df$LenderYield),2)), size = 4, color = "#ff7f0e")
```


The lender yield is the interest rate minus the expected fee payments. It is the most important input into any return calculation. As we can see, lender yield is 0.1827 on average. Most of the laons range from 0.1242 to  0.2400.


# Univariate Analysis

### What is the structure of your dataset?

```{r echo=FALSE, Summary}
summary(df)
```

### What is/are the main feature(s) of interest in your dataset?


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!