Exploratory Data Analysis of Loan Data from Prosper
_Barbara Stempien_
========================================================

Prosper Marketplace is America's first peer-to-peer lending marketplace, with
over \$7 billion in funded loans. Borrowers request personal loans on Prosper
and investors (individual or institutional) can fund anywhere from \$2,000 to 
\$35,000 per loan request. Investors can consider borrowers’ credit scores, 
ratings, and histories and the category of the loan. Prosper handles the 
servicing of the loan and collects and distributes borrower payments and 
interest back to the loan investors.

Prosper verifies borrowers' identities and select personal data before 
funding loans and manages all stages of loan servicing. Prosper's unsecured
personal loans are fully amortized over a period of three or five years, with
no pre-payment penalties. Prosper generates revenue by collecting a one-time
fee on funded loans from borrowers and assessing an annual loan servicing fee
to investors.

Prosper publishes performance statistics on its website and all market data
is available to the public for analysis. Prosper loan data set contains 
113,937 loans with 81 variables on each loan, 
including loan amount, borrower rate (or interest rate), current loan status, 
borrower income, borrower employment status, borrower credit history, and 
the latest payment information.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='plots/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
# Load packages
library(ggplot2)
library(repr)
library(ggthemes)
library(dplyr)
library(maps)
library(mapdata)
library(openintro)
library(waffle)
library(ggcorrplot)
library(scales)
library(ggpubr)
library(gridExtra)
library(GGally)
library(knitr)
```

```{r}
# load the prosper loan dataset
loan_dataset <- read.csv('data/prosperLoanData.csv', na.strings = c("", "NA"))
```

```{r}

# subset the dataframe to the following columns
subset <- c("LoanStatus",
            "ListingCreationDate",
            "ClosedDate",
            "ListingCategory..numeric.",
            "Term",
            "BorrowerRate",
            "LoanOriginalAmount",
            "MonthlyLoanPayment",
            "BorrowerState",
            "IsBorrowerHomeowner",  
            "Occupation",
            "EmploymentStatus",
            "StatedMonthlyIncome",
            "DebtToIncomeRatio",
            "ProsperRating..Alpha.",
            "ProsperScore",           
            "OpenCreditLines",
            "TotalCreditLinespast7years",
            "OpenRevolvingAccounts",
            "OpenRevolvingMonthlyPayment",
            "CurrentDelinquencies",
            "AmountDelinquent",
            "DelinquenciesLast7Years",
            "PublicRecordsLast10Years",
            "RevolvingCreditBalance",
            "BankcardUtilization",
            "LenderYield",
            "Investors"
           )

# subset dataframe
df <- loan_dataset[subset]

# replace NAs with 0
df[is.na(df)] <- 0
```
``` {r ,Clean and Tidy dataset}

# convert Loan Status to factor with 7 levels
# https://stackoverflow.com/a/19410249/7382214
levels(df$LoanStatus) <- list(Current = "Current",
                              Completed = "Completed",
                              "Charged Off" = "Chargedoff",
                              Defaulted = "Defaulted",
                              "Past Due" = c("Past Due (>120 days)",
                                        "Past Due (1-15 days)",
                                        "Past Due (16-30 days)",
                                        "Past Due (31-60 days)",
                                        "Past Due (61-90 days)",
                                        "Past Due (91-120 days)"),
                              "Final Payment" = "FinalPaymentInProgress",
                              Cancelled = "Cancelled")

# convert Listing Creation Date variable to date format
df$ListingCreationDate <- as.Date(df$ListingCreationDate)

# convert Term to ordered factor with 3 levels
df$Term <- factor(df$Term, levels=c(12,36,60), ordered=TRUE)

# convert Closed Date variable to date format
df$ClosedDate <- as.Date(df$ClosedDate)

# convert Prosper Rating to ordered factor with 8 levels
df$ProsperRating..Alpha. <- factor(df$ProsperRating..Alpha.,
                                   levels=c("HR", "E", "D", "C","B","A","AA"),
                                   ordered=TRUE)

# convert Prosper Score to ordered factor with 12 levels
df$ProsperScore <- factor(df$ProsperScore,
                          levels=c(0,1,2,3,4,5,6,7,8,9,10,11),
                          ordered=TRUE)

# add column with Listing Category labels based on the documentation
df$ListingCategory <- factor(df$ListingCategory..numeric., 
                             labels=c('Not Available',
                                      'Debt Consolidation',
                                      'Home Improvement',
                                      'Business',
                                      'Personal Loan',
                                      'Student Use',
                                      'Auto',
                                      'Other',
                                      'Baby&Adoption',
                                      'Boat',
                                      'Cosmetic Procedure',
                                      'Engagement Ring',
                                      'Green Loans',
                                      'Household Expenses',
                                      'Large Purchases',
                                      'Medical/Dental',
                                      'Motorcycle',
                                      'RV',
                                      'Taxes',
                                      'Vacation',
                                      'Wedding Loans'))

# move ListingCategory column next to the ListingCategory..numeric.
df <- df[c(1:4,29,5:28)]

# convert State abbreviation to full name
df$BorrowerState <- abbr2state(df$BorrowerState)

# convert Is Borrower Homeowner to logical type
df$IsBorrowerHomeowner <- as.logical(df$IsBorrowerHomeowner)

# replace NA in Occupation with Not Available
levels(df$Occupation)<-c(levels(df$Occupation),"Not Available") 
df$Occupation[is.na(df$Occupation)] <- "Not Available"

# replace NA in Employment Status with Not Available
df$EmploymentStatus[is.na(df$EmploymentStatus)] <- "Not available"

```

# Univariate Plots Section

> We will begin the analysis by looking at the data contained in the data set.

```{r}
head(loan_dataset)
```

```{r}
summary(df)
```

First, we want to see when and how many listings were created.

```{r}
# Group Listings by year and month
creation_date <- df %>% count(df$ListingCreationDate) %>%
                setNames(.,c("date", "count")) %>%
                mutate(month = as.Date(cut(date, breaks = "month")))  %>%
                group_by(month) %>% 
                summarise(total = sum(count)) 

# Plot Listing Creation Date
ggplot(creation_date, aes(x=month, y=total)) +
      geom_line(color = "#3F5D7D", size = 0.75) + 
      geom_point(color = "#3F5D7D", size = 3) +
      scale_x_date(date_labels = "%b-%Y", date_breaks="4 months") +
      labs(title = "When was the listing created?") + 
      scale_y_continuous(limits = c(0,6000), breaks = seq(0, 6000, 1000)) +
      theme(plot.title = element_text(size = 18),
            axis.title.x = element_blank(),
            axis.title.y = element_blank(),
            axis.text.y = element_text(size=10),
            axis.text.x = element_text(angle = 45, hjust = 1, size=10),
            panel.grid.major.y = element_line(color="#d3d3d3", 
                                              linetype = "dashed", 
                                              size=0.5))

```

From the graph above, we can see that the number of loans in Prosper has
increased over time. In 2008 there was a collapse in the number of listings,
however, since 2009, the number has been constantly growing.

```{r}
# Loan Original Amount statistics
summary(df$LoanOriginalAmount)

# plot Loan Original Amount
ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(binwidth = 1000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1000, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(breaks = seq(0, 36000, 1000), 
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,17500), 
                     breaks = seq(0, 17500, 2500), 
                     expand = c(0,0)) +
  labs(title="How high is the original loan amount (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(LoanOriginalAmount)),
            color = "#ff7f0e", 
            linetype = "dashed", 
            size = 1) + 
  annotate("text", x = 10000, y = 17000, 
           label = paste("Mean: $", round(mean(df$LoanOriginalAmount))), 
           size = 4, 
           color = "#ff7f0e")
```
 
Another interesting feature is the amount of loans. In the above graph we can
see that most loans are granted for amounts from \$4,000 to \$12,000. 
The average loan amount is \$8337.


```{r}
# Term
term <- df %>% count(df$Term) %>%
                    setNames(.,c("term", "count"))

# plot term
waffle(c("12 months" = term$count[term$term == 12] / 500, 
         "36 months" = term$count[term$term == 36] / 500,
         "60 months" = term$count[term$term == 60] / 500),
       rows = 10, 
       colors = c("#d3d3d3", "#a9bacc", "#3F5D7D"), 
       size = 0.5, 
       pad = 1, 
       legend_pos = "bottom", 
       title = "How long is a loan term?",
       xlab = "1 square = 500 customers")
```

The vast majority of loans is granted for 36 months. Only a fraction of all
loans is granted for 12 months.


```{r}

# Listing Category
listing_category <- df %>% count(df$ListingCategory) %>%
                    setNames(.,c("category", "count"))

# plot listing category
ggplot(listing_category, aes(x =  reorder(category, -count), y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  labs(title = "What is the listing's category?") + 
  scale_y_continuous(limits = c(0,65000), 
                     breaks = seq(0, 65000, 5000), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

Consolidation of debts is the most common reason for a loan. 
We do not have information about the category for a large
number of loans - 16,000. The second most common category is Other,
followed by Home Improvement and Business.

```{r}

# Loan Status 
loan_status <- df %>% count(df$LoanStatus) %>%
                setNames(.,c("status", "count")) %>%
                arrange(desc(count))

# plot loan status
ggplot(loan_status, aes(x = status, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.25) +
  labs(title = "What is the loan status?") + 
  scale_y_continuous(limits = c(0,60000),
                     breaks = seq(0, 55000, 5000),
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3",
                                          linetype = "dashed", 
                                          size=0.5)) + 
  rotate()

```

Most of the loans in the dataset are currently active (56,000). 
Over 38,000, was paid off. Around 16 thousand were charged off or defaulted.

```{r}
# Borrower Rate
summary(df$BorrowerRate)
        
borrower_rate <- df %>% count(df$BorrowerRate) %>%
                    set_colnames(c("rate", "count"))

ggplot(aes(x = BorrowerRate * 100), data = df) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, 
           aes(y=..count.., label=..count..,), 
           geom="text", vjust=-.5, 
           size=3) +
  scale_x_continuous(breaks = seq(0, 50, 1), 
                     labels=dollar_format(prefix="", suffix="%")) +
  scale_y_continuous(limits = c(0,6500), 
                     breaks = seq(0, 6500, 1000), 
                     expand = c(0,0)) +
  labs(title="What is the borrower's rate?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(BorrowerRate) * 100),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 22, y = 6250, 
           label = paste("Mean:", round(mean(df$BorrowerRate * 100),2), "%"), 
           size = 4, 
           color = "#ff7f0e")
```

As for the borrower's rate, it is 0.1928 on average. Although a considerable
amount of loans is granted at the borrower's rate of 0.14, 0.15, 0.18,
a lot of loans are granted at a much higher rate, for instance 
0.32 (5914 loans).

```{r}
# Monthly Loan Payment statistics

summary(df$MonthlyLoanPayment)

# plot Monthly Loan Payment 
ggplot(aes(x = MonthlyLoanPayment), data = df) +
  geom_histogram(binwidth = 100, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 100, 
           aes(y=..count.., label=..count..,), 
           geom="text", vjust=-.5, 
           size=3) +
  scale_x_continuous(breaks = seq(0, 2300, 100), 
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,30000), 
                     breaks = seq(0, 30000, 5000), 
                     expand = c(0,0)) +
  labs(title="How high is the monthly loan payment (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(MonthlyLoanPayment)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 400, y = 29000, 
           label = paste("Mean: $", round(mean(df$MonthlyLoanPayment))), 
           size = 4, color = "#ff7f0e")
```

The average loan installment is \$272. Most loan installments are
in the \$130 - \$370 range.

```{r}
# group loans by borrower's state
loans <- df %>% count(df$BorrowerState) %>%
                setNames(.,c("region", "count"))
loans$region <- tolower(loans$region)

# get polygon data for states
states <- map_data("state")

# merge datasets
loans_states <- inner_join(states, loans, by = "region")

# plot borrowers by state
ggplot(data = loans_states, mapping = aes(x = long, y = lat, group = group)) + 
  coord_fixed(1.3) + 
  geom_polygon(color = "#000000", fill = "#d3d3d3") + 
  geom_polygon(data = loans_states, aes(fill = count), color = "#ffffff") +
  geom_polygon(color = "#000000", fill = NA) +
  labs(title = "Where does the borrower live?") + 
  theme_bw() +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank()) + 
  scale_fill_gradient(trans = "log10", low = "#d3d3d3", high = "#3F5D7D")
```

The largest number of loans is granted to borrowers living in the California.

```{r}
# IsBorrowerHomeowner variable
homeowner <- df %>% count(df$IsBorrowerHomeowner) %>%
                setNames(.,c("owner", "count"))

# plot home ownership
waffle(c(No = homeowner$count[homeowner$owner == "FALSE"] / 1000, 
         Yes = homeowner$count[homeowner$owner == "TRUE"] / 1000), 
       rows = 6, 
       colors = c("#d3d3d3","#3F5D7D"), 
       size = 0.5, pad = 1, 
       legend_pos = "bottom", 
       title = "Is Borrower a Homeowner?",
       xlab = "1 square = 1000 customers")
```

The share of homeowners in the total number of borrowers is around 50%.

```{r}
# Employment Status
employment <- df %>% count(df$EmploymentStatus) %>%
                    setNames(.,c("employment", "count")) %>%
                    arrange(desc(count))

# plot Employment Status
ggplot(employment, aes(x = reorder(employment, count), y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.5) +
  labs(title = "What is the borrower's employment status?") + 
  scale_y_continuous(limits = c(0,75000), 
                     breaks = seq(0, 70000, 5000), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

The vast majority of borrowers are currently employed. Only a fraction
of borrowers do not have a job or work part-time.

```{r}
# Occupation
occupation <- df %>% count(df$Occupation) %>%
                    setNames(.,c("occupation", "count")) %>%
                    arrange(desc(count))

# plot occupation
ggplot(occupation[1:34,], aes(x = reorder(occupation, -count), y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  labs(title = "What is the borrower's occupation?") + 
  scale_y_continuous(limits = c(0,35000), 
                     breaks = seq(0, 35000, 5000), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, 
                                   hjust = 1, 
                                   size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))

ggplot(occupation[35:67,], 
       aes(x = reorder(occupation, -count), y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), vjust = -0.5) +
  scale_y_continuous(limits = c(0,35000), 
                     breaks = seq(0, 35000, 5000), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

Unfortunately, for the majority of loans, there is no information 
about the borrower's profession. A significant part of borrowers are 
professionals, programmers or executives.

```{r}
# Stated Monthly Income
summary(df$StatedMonthlyIncome)

# plot Stated Monthly Income
ggplot(aes(x = StatedMonthlyIncome), data = df) +
  geom_histogram(binwidth = 1000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1000, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(-999, quantile(df$StatedMonthlyIncome, 0.99)), 
                     breaks = seq(0, 20000, 1000),
                     expand = c(0,0),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,20000), 
                     breaks = seq(0, 20000, 2500), 
                     expand = c(0,0)) +
  labs(title="What income was delcared by the borrower (monthly, USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(StatedMonthlyIncome)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", 
           x = 6500,
           y = 19500, 
           label = paste("Mean: $", round(mean(df$StatedMonthlyIncome))), 
           size = 4, 
           color = "#ff7f0e")
```

The average monthly income of a borrower is \$5608. The vast majority 
of borrowers earn in the period between \$3200 - \$6825.

```{r}
# Debt To Income Ratio
summary(df$DebtToIncomeRatio)

# plot debt to income ratio
ggplot(aes(x = DebtToIncomeRatio), data = df) +
  geom_histogram(binwidth = 0.05, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 0.05, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(-0.05, quantile(df$DebtToIncomeRatio, 0.99)), 
                     breaks = seq(0, 2, 0.05), 
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(0,20000), 
                     breaks = seq(0, 20000, 2500), 
                     expand = c(0,0)) +
  labs(title="What is the borrower's debt to income ratio?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(DebtToIncomeRatio)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 0.29, y = 19000, 
           label = paste("Mean:", round(mean(df$DebtToIncomeRatio),2)), 
           size = 4, 
           color = "#ff7f0e")
```

It is interesting to compare income to debt. As we can see in the plot above, 
the average debt to income ratio is 0.2552 and the vast majority of borrowers 
are in the range of 0.1300 - 0.3100.

```{r}
# Delinquencies
limit <- quantile(df$DelinquenciesLast7Years, 0.99)

ggplot(data = df, aes(x = DelinquenciesLast7Years)) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(-1, limit), breaks = seq(0, limit, 1)) +
  scale_y_continuous(limits = c(0,80000), 
                     breaks = seq(0, 80000, 5000), 
                     expand = c(0,0)) +
  labs(title = "How many delinquencies did borrower have in past 7 years??") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(DelinquenciesLast7Years)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", x = 6, y = 75000, 
           label = paste("Mean:", round(mean(df$DelinquenciesLast7Years))), 
           size = 4, 
           color = "#ff7f0e")
```

Another important aspect when assessing credit risk is the number
of delinquencies that the borrower had in recent years. As we can see in
the plot above, most borrowers did not have any delinquencies 
in the last 7 years.

```{r}
# Public Records
limit <- quantile(df$PublicRecordsLast10Years, 0.99)

ggplot(data = df, aes(x = PublicRecordsLast10Years)) +
  geom_histogram(binwidth = 1, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 1, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(-1, limit), 
                     breaks = seq(0, limit, 1)) +
  scale_y_continuous(limits = c(0,90000), 
                     breaks = seq(0, 90000, 5000), 
                     expand = c(0,0)) +
  labs(title = "How many public records did borrower have in past 10 years??") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(PublicRecordsLast10Years)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", 
           x = 6, 
           y = 75000, 
           label = paste("Mean:", round(mean(df$PublicRecordsLast10Years))), 
           size = 4, 
           color = "#ff7f0e")
```

We see similar situations when it comes to public records
from the last 10 years. For a significant number of borrowers, it is 0.

```{r}
# Bank card Utilization
summary(df$BankcardUtilization)

ggplot(aes(x = BankcardUtilization * 100), data = df) +
  geom_histogram(binwidth = 5, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 5, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(-3,125), 
                     breaks = seq(0, 125, 5), 
                     expand = c(0,0), 
                     labels=dollar_format(prefix="", suffix="%")) +
  scale_y_continuous(limits = c(0,18000), 
                     breaks = seq(0, 18000, 2500), 
                     expand = c(0,0)) +
  labs(title="What is the borrower's bankcard utilization?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(BankcardUtilization * 100)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text",
           x = 58, 
           y = 16000, 
           label = paste("Mean:", 
                         round(mean(df$BankcardUtilization * 100),2), "%"), 
           size = 4, 
           color = "#ff7f0e")
```

Utilization of a credit card is also an important factor in risk assessment. 
As we can see in the plot above, the average credit card utilization is 52%, 
and most borrowers used between 23% and 82% of the available resources
on the credit card.

```{r}
# Revolving Credit Balance
summary(df$RevolvingCreditBalance)

# Revolving Credit Balance
ggplot(aes(x = RevolvingCreditBalance), data = df) +
  geom_histogram(binwidth = 5000, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 5000, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits = c(0, quantile(df$RevolvingCreditBalance, 0.99)), 
                     breaks = seq(0, 150000, 5000),
                     expand = c(0,0),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0,30000), 
                     breaks = seq(0, 30000, 2500), 
                     expand = c(0,0)) +
  labs(title="What is the amount of borrower's 
       revolving credit balance (USD)?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(RevolvingCreditBalance)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", 
           x = 24000, 
           y = 28000, 
           label = paste("Mean: $", round(mean(df$RevolvingCreditBalance))), 
           size = 4, 
           color = "#ff7f0e")
```

Another value that can help us analyze the financial status of the borrower 
is revelving credit balance. As we can see, the average value is \$16,424,
and most borrowers are in the \$2091 - \$18254 range.


```{r}
# Prosper Rating
prosper_rating <- df %>% count(df$ProsperRating..Alpha.) %>%
                    setNames(.,c("rating", "count")) %>%
                    arrange(desc(count))

# Prosper Rating
ggplot(prosper_rating, aes(x = reorder(rating, count), y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.5) +
  labs(title = "What is the borrower's prosper rating?") + 
  scale_y_continuous(limits = c(0,31000), 
                     breaks = seq(0, 31000, 2500), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

Based on these data, and several other factors, Prosper assessed
the credit risk associated with each borrower. As we can see on the graph above,
the most common risk is C, then B. For a large number of borrowers, 
we do not have any information about the risk, as this information was
not collected before 2009.

```{r}
# Number of investors
summary(df$Investors)

ggplot(aes(x = Investors), data = df) +
  geom_density(color = "#ffffff", fill = "#3F5D7D") + 
  scale_x_continuous(limits = c(0,650), breaks = seq(0, 650, 25)) +
  labs(title="What is the number of investors?") + 
  xlab('Number of investors') +
  ylab('Density') +
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(Investors)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", 
           x = 110, 
           y = 0.015, 
           label = paste("Mean:", round(mean(df$Investors),0)), 
           size = 4, 
           color = "#ff7f0e")
```

The average number of investors per loan is 80. Most of the loans have 
from 1 to 115 investors. The highest number of investors per loan was 1189.

```{r}
# Lender Yield
summary(df$LenderYield)

ggplot(aes(x = LenderYield), data = df) +
  geom_histogram(binwidth = 0.025, color = "#ffffff", fill = "#3F5D7D") +
  stat_bin(binwidth = 0.025, 
           aes(y=..count.., label=..count..,), 
           geom="text", 
           vjust=-.5, 
           size=3) +
  scale_x_continuous(limits=c(0,0.37),breaks = seq(0, 0.35, 0.025)) +
  scale_y_continuous(breaks = seq(0, 14000, 2000)) +
  labs(title="What is the lender yield?") + 
  theme(plot.title = element_text(size = 18, vjust = 5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 45, hjust = 1, size=10),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  geom_vline(aes(xintercept = mean(LenderYield)),
            color = "#ff7f0e", linetype = "dashed", size = 1) + 
  annotate("text", 
           x = 0.205, 
           y = 14000, 
           label = paste("Mean:", round(mean(df$LenderYield),2)), 
           size = 4, 
           color = "#ff7f0e")
```


The lender yield is the interest rate minus the expected fee payments. 
It is the most important input into any return calculation. As we can see, 
lender yield is 0.1827 on average. Most of the laons range 
from 0.1242 to  0.2400.


# Univariate Analysis

### What is the structure of your dataset?

```{r echo=FALSE, Summary}
summary(df)
```


Prosper loan data set contains 113,937 loans with 81 variables on each loan. 
I have selected 29 variables from the original dataset for my analysis.


### What is/are the main feature(s) of interest in your dataset?

The most interesting for me is to understand what is the profile of typical 
peer-to-peer borrower, who fails to pay off the loan, What are 
the demographics and credit characteristics of the borrower 
who defaults is past due on the credit.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Features like lender yield or number of investors might be helpful to
understand which features of the borrower affect the increase of investors'
interest, and therefore are desirable, and which are not. 

### Did you create any new variables from existing variables in the dataset?

Yes, I created new Listing Category variable, with textual values, 
based on the original ListingCategory..numeric. variable.           

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I have :
* grouped various Past Due statuses of the Loan Status variable into one
Past Due status, and converted this variable into factor with 7 levels;
* converted Listing Creation Date variable to date format;
* converted Term to ordered factor with 3 levels;
* converted Closed Date variable to date format;
* converted Prosper Rating to ordered factor with 8 levels;
* converted Prosper Score to ordered factor with 12 levels;
* converted State abbreviation to full name and added as new variable;
* converted Is Borrower Homeowner to logical type;
* replaced NA in Occupation with Not Available;
* replaced NA in Employment Status with Not Available;

These changes have been performed to ease plotting.
 
While building different plots, I have also grouped variables and subset
dataset whenever needed.


# Bivariate Plots Section

```{r}
# plot correlation matrix
ggcorr(df, label = TRUE, label_size = 3, 
       hjust = 0.8, size = 2.5, color = "black", layout.exp = 2)
```

Correlogram shows that there is a strong correlation (>= 0.7) between:

- Open Revolving Monthly Payment and Revolving Credit Balance;
- Open Credit Lines and Open Revolving Accounts;
- Loan Orginal Amount and Laon Monthly Payment;
- Borrower Rate and Lender Yield;

These correlations are not surprising. In my analysis, I plan to focus on 
the variables that impact borrowers ability to pay off the credit, therefore
Loan Orginal Amount and Laon Monthly Payment will definitely be investigated.


```{r echo=FALSE, Bivariate_Plots}
# average prosper score by loan status
score_status <- df[c(1,17)]
score_status$ProsperScore <- as.numeric(levels(
  score_status$ProsperScore))[score_status$ProsperScore]
df.avg_score_status <- score_status %>%
    group_by(LoanStatus) %>%
    summarise(AvgProsperScore = mean(ProsperScore,na.rm = TRUE), 
              Count = n()) %>%
    arrange(AvgProsperScore)

df.avg_score_status

# plot average prosper score by loan status
ggplot(df.avg_score_status, aes(x = reorder(LoanStatus, -AvgProsperScore) , 
                                y = AvgProsperScore)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgProsperScore,2)), hjust = -0.5) +
  labs(title = "What is the average prosper score by loan status?") + 
  scale_y_continuous(limits = c(0,7), breaks = c(0,1,2,3,4,5,6,7)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

Looking at the above plot we can see that the average prosper score is very low
for loans that were defaulted or charged off. Interestingly, the prosper score
is also low for loans that are already completed and rather high for loans that
are past due. The highest prosper score have current loans and those awaiting
a final payment.

```{r}
# average prosper loan amount by loan status
amount_status <- df[c(1,8)]
df.avg_amount_status <- amount_status %>%
    group_by(LoanStatus) %>%
    summarise(AvgLoanOriginalAmount = mean(LoanOriginalAmount,na.rm = TRUE), 
              Count = n()) %>%
    arrange(AvgLoanOriginalAmount)

df.avg_amount_status

# plot average loan amount by loan status
ggplot(df.avg_amount_status, 
       aes(x = reorder(LoanStatus, -AvgLoanOriginalAmount) , 
                                 y = AvgLoanOriginalAmount)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgLoanOriginalAmount,0)), hjust = -0.5) +
  labs(title = "What is the average loan original amount by loan status?") + 
  scale_y_continuous(limits = c(0,11000), breaks = seq(0, 11000, 1000)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

As we can see on the above plot, the average loan amount for charged off and
defaulted loans is much lower than for current loans. Interestingly, 
the average loan amount is aslo lower for completed loans. 
The average amount of loan for current loans is \$10,361.

```{r}
# average prosper loan amount by prosper score
amount_score <- df[c(17,8)]
df.avg_amount_score <- amount_score %>%
    group_by(ProsperScore) %>%
    summarise(AvgLoanOriginalAmount = mean(LoanOriginalAmount,na.rm = TRUE), 
              Count = n()) %>%
    arrange(ProsperScore)

df.avg_amount_score

# plot average loan amount by prosper score
ggplot(df.avg_amount_score, aes(x = ProsperScore, y = AvgLoanOriginalAmount)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgLoanOriginalAmount,0)), hjust = -0.5) +
  labs(title = "What is the average loan original amount by prosper score?") + 
#   scale_y_continuous(limits = c(0,11000), breaks = seq(0, 11000, 1000)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

Another interesting observation we can make is about the relation of the loan
amount to the borrower's score. In the above plot (N/A values are removed) 
we can see that the higher amounts of the loan are granted to borrowers
with the highest rating. This is in line with common sense - larger loans 
are granted to a more reliable borrower, while smaller loans to those 
with a greater risk of non-repayment.


```{r}
# monthly loan payment by loan status
monthly_status <- df[c(1,9)]
df.avg_monthly_status <- monthly_status %>%
    group_by(LoanStatus) %>%
    summarise(AvgMonthlyPayment = mean(MonthlyLoanPayment,na.rm = TRUE), 
              Count = n()) %>%
    arrange(AvgMonthlyPayment)

df.avg_monthly_status

# plot monthly loan payment by loan status
ggplot(df.avg_monthly_status, aes(x = reorder(LoanStatus, -AvgMonthlyPayment), 
                                  y = AvgMonthlyPayment)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgMonthlyPayment,0)), hjust = -0.5) +
  labs(title = "What is the average monthly loan payment by loan status?") + 
  scale_y_continuous(limits = c(0,400), 
                     breaks = seq(0, 400, 50), 
                     labels=dollar_format(prefix="$")) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

As we can see in the above plot, the average monthly payment is the highest
for current loans, followed shortly by the loans awaiting final payment. 
Past due loans are in the third position. We can also see that the average 
monthly loan payment for completed loans is one of the lowest. This is as
expected - smaller loans are easier to pay back while larger loans, 
that more time and effort.


```{r}
# Stated Monthly Income by loan status
monthly_income_status <- df[c(1,14)]
df.avg_monthly_income_status <- monthly_income_status %>%
    group_by(LoanStatus) %>%
    summarise(AvgMonthlyPayment = mean(StatedMonthlyIncome,na.rm = TRUE), 
              Count = n()) %>%
    arrange(AvgMonthlyPayment)

df.avg_monthly_income_status

# plot Stated Monthly Income by loan status
ggplot(df.avg_monthly_income_status, 
       aes(x = reorder(LoanStatus, -AvgMonthlyPayment),
           y = AvgMonthlyPayment)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgMonthlyPayment,0)), hjust = -0.5) +
  labs(title = "What is the average monthly income by loan status?") + 
  scale_y_continuous(limits = c(0, 6500), 
                     breaks = seq(0, 6500, 500), 
                     labels=dollar_format(prefix="$")) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

As we can see in the above plot, the average stated monthly income is
the highest for loans awaiting final payment, followed shortly by 
the current loans. Past due loans are in the third position, close to
Completed loans. Low income is typical for canceled, defaulted 
and charged-off loans.

```{r}
# average debt to income ratio by loan status
debt_status <- df[c(1,15)]
df.avg_debt_status <- debt_status %>%
    group_by(LoanStatus) %>%
    summarise(AvgDebtToIncomeRatio = mean(DebtToIncomeRatio,na.rm = TRUE), 
              Count = n()) %>%
    arrange(AvgDebtToIncomeRatio)

# plot average debt to income by loan status
ggplot(df.avg_debt_status, aes(x = reorder(LoanStatus, -AvgDebtToIncomeRatio) , 
                               y = AvgDebtToIncomeRatio)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = round(AvgDebtToIncomeRatio,2)), hjust = -0.5) +
  labs(title = "What is the average debt to income ratio by loan status?") + 
#   scale_y_continuous(limits = c(0,7), breaks = c(0,1,2,3,4,5,6,7)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) +
  rotate()
```

It's not surprising to see that the average debt to income ratio is the highest
for defaulted, past due and charged off loans. We can also see it is 
significantly lower for loans awaiting the final payment and completed.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset? 
Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

An interesting relationship I have observer is that prosper score and 
loan status are not related as we could expect. Quite a high number of 
loans with high prosper score is past due. Moreover, the average prospers
score for completed loans is much lower than for past due loans.

Another interesting insight is that the average loan amount for borrowers 
with the lowest score - 0 - is higher than the average loan amount for 
borrowers with score 1 and score 2. This is surprising, as the risk related 
to investing in a borrower with score 0 is much higher than investing in a 
borrower with score 1 or 2.

### What was the strongest relationship you found?

The strongest relationship I have found was between debt to income ratio 
and loan status. Borrowers with the high debt to income ratio were more 
often defaulting, past due or charged off on their loan. 

# Multivariate Plots Section

Let's start from looking at the relation of employment status, debt to 
income ratio and loan status.

```{r}
# plot employmnt status, debt to income ratio and loan status
ggplot(aes(EmploymentStatus, DebtToIncomeRatio), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_y_continuous(breaks = seq(0, 10, 0.5)) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Employment Status, Debt To Income Ratio and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

As we can see in the above plot, only a small fraction of the loans belong to
the unemployed borrowers. We can see most of the current loans are in 
the employed and other buckets. Unfortunately, most of the defaulted loans 
are in the not available bucket. Debt to income ratio is below 1 for most 
of the loans. We can see however a small pick at the ratio of 10.  To get 
a better understanding of these relations, let's zoom in to loans with debt
to income ratio below the 99 quantile.

```{r}
ggplot(aes(EmploymentStatus, DebtToIncomeRatio), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_y_continuous(limits = c(0.0, quantile(df$DebtToIncomeRatio, 0.99)),
                     breaks = seq(0, 1.5, 0.1)) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Employment Status, Debt To Income Ratio and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```


Looking at this plot, we can see that there is significantly less Charged off 
loans in the employed bucket. We can see them however in the full-time bucket.
We can also see that defaulted loans are spread rather equally when it comes 
to the borrower's debt to income ratio. This is rather surprising, as we would
expect borrowers with the highest debt to income ratio the default on their 
loans more frequently.


Now, let's see how monthly loan payment relates to employment status 
and loan status.

```{r}
ggplot(aes(EmploymentStatus, MonthlyLoanPayment), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_y_continuous(labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Employment Status, Monthly Loan Payment and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

As we can see in the above plot, most of the loans have monthly payment 
below 1500\$. The self-employed bucket is quite interesting, as we can 
see that loans closer to 1000$ then do be charged off or defaulted. 
To get a better understanding of these relations, let's zoom in to 
loans with monthly payment below the 99 quantile.

```{r}
ggplot(aes(EmploymentStatus, MonthlyLoanPayment), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_y_continuous(limits = c(0.0, quantile(df$MonthlyLoanPayment, 0.99)),
                     breaks = seq(0, 1000, 50),
                     labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Employment Status, Monthly Loan Payment and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

Monthly loan payment does not seem to have a big impact on the loan status
for employed borrowers. We can see only a slight increase in the number 
of charged off and defaulted loans for the full time and not employed
borrowers. For self-employed borrowers, we can see that the number of 
charged-off or defaulted loans with monthly payment above the $600 is 
higher than for lower monthly payments.

Another interesting observation we can make is that borrowers who are not 
employed, retired or work part-time, then to be offered loans with a monthly
payment below \$400. Majority of loans granted to self-employed borrowers 
have monthly payment below \$600. At the same time, employees and full-time
workers get loans with monthly payment below $900.

Let's now look at the relation of Stated Monthly Income, Monthly Loan Payment
and Loan Status.

```{r}
ggplot(aes(StatedMonthlyIncome, MonthlyLoanPayment), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_x_continuous(labels=dollar_format(prefix="$")) +
  scale_y_continuous(labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Stated Monthly Income, Monthly Loan Payment and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))

```

As we can see, we have some outliers in the Stated Monthly Income, 
let's zoom in to 99 quantile.

```{r}
ggplot(aes(StatedMonthlyIncome, MonthlyLoanPayment), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_x_continuous(limits = c(0.0, quantile(df$StatedMonthlyIncome, 0.99)),
                     breaks = seq(0, 100000, 1000),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0.0, quantile(df$MonthlyLoanPayment, 0.99)),
                     breaks = seq(0, 1500, 100),
                     labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Stated Monthly Income, Monthly Loan Payment and 
Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

As we can see for the majority of loans monthly loan payment is below \$700 
while borrowers stated income is below \$10,000. It is interesting to see a lot
of charged-off and defaulted loans in the lower left corner of the plot - small
monthly payment but a small income. It seems that borrowers with higher income
can pay off the loan even if the monthly payment is larger, while borrowers 
with the low-income struggle to pay off the loan even if the monthly payment
is low.

```{r}
ggplot(aes(ProsperRating..Alpha., LoanOriginalAmount), data=df) +
  geom_boxplot(aes(fill=LoanStatus),alpha=1) +
  scale_y_continuous(breaks = seq(0, 50000, 2500),
                     labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = 
         "Prosper Rating, Loan Original Amount and Loan Status relation") +
  ylab("Loan Original Amount") +
  xlab("Prosper Rating") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

In this plot, we can look closer at the prosper rating, loan original amount 
and loan status. We can see that most of the completed loans were loans with 
a small amount.  Most of the loans with a high amount are still active. 
In this plot, we can also clearly see a relation between prosper score and 
the amount of loan. Borrowers with a low score are granted smaller loans, 
while those with good rating bigger loans. High Rish and E category borrowers
are almost never granted a loan above the \$10,000, while we can see that 
some of the AA, A and B category borrowers have loans above the \$25,000.

Lastly, let's have a look if there is any significant relation between Prosper
Rating, Listing Category and Loan Status.

```{r}
ggplot(aes(ProsperRating..Alpha., ListingCategory), data=df)+
  geom_jitter(aes(color=LoanStatus),alpha=1)+
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Prosper Rating, Listing Category and Loan Status relation") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

It seems that there is no relation between the listing category and
loan status. There is no category that would have significantly more 
defaulted or charged-off loans.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest? 
Were there any interesting or surprising interactions between features?

I noticed that only a small fraction of the loans belong to the unemployed 
borrowers, most of the current loans belong to the employed, full-time
borrowers. At first glance, we could see that employed borrowers have 
a significantly less Charged off loans, however, these loans appear 
frequently for full-time working borrowers. Unfortunately, for most of 
the defaulted loans, we do not have information on the borrower's employment
status, which limits our investigation.

Another interesting finding is that the debt to income ratio is below 1
for most of the loans.  We could also see that defaulted loans are spread 
rather equally when it comes to the borrower's debt to income ratio. 
This is rather surprising, as we would expect borrowers with the highest 
debt to income ratio the default on their loans more frequently.

When it comes to how monthly loan payment relates to employment status and
loan status, we could see that monthly loan payment does not seem to have 
a big impact on the loan status for employed borrowers. We could see only 
a slight increase in the number of charged off and defaulted loans for the
full time and not employed borrowers. However, for self-employed borrowers,
we could see that the number of charged-off or defaulted loans with monthly
payment above the $600 is higher than for lower monthly payments.

Another interesting observation we made is that borrowers who are not 
employed, retired or work part-time, then to be offered loans with a monthly
payment below \$400. Majority of loans granted to self-employed borrowers
have monthly payment below \$600. At the same time, employees and full-time
workers get loans with monthly payment below $900.

The majority of loans monthly loan payment is below \$700 while borrowers
stated income is below \$10,000. It was interesting to see a lot of 
charged-off and defaulted loans for borrowers with a small monthly 
payment and a small income. It seems that borrowers with higher income 
can pay off the loan even if the monthly payment is larger, while borrowers
with the low-income struggle to pay off the loan even if the monthly 
payment is low.

We could also see that most of the completed loans were loans with a small
amount.  Most of the loans with a high amount are still active. We could 
also clearly see a relation between prosper score and the amount of loan.
Borrowers with a low score are granted smaller loans, while those with 
good rating bigger loans. High Rish and E category borrowers are almost
never granted a loan above the \$10,000, while we can see that some 
of the AA, A and B category borrowers have loans above the \$25,000.

Surprisingly, we did not find any relation between the listing category 
and loan status. There is no category that would have significantly more 
defaulted or charged-off loans.



------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Loan Status 
loan_status <- df %>% count(df$LoanStatus) %>%
                setNames(.,c("status", "count")) %>%
                arrange(desc(count))

# plot loan status
ggplot(loan_status, aes(x = status, y = count)) +
  geom_bar(fill = "#3F5D7D", stat = "identity") +
  geom_text(aes(label = count), hjust = -0.25) +
  labs(title = "What is the loan status?") + 
  xlab("Loan Status") +
  ylab("Count") +
  scale_y_continuous(limits = c(0,60000), 
                     breaks = seq(0, 55000, 5000), 
                     expand = c(0,0)) +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5)) + 
  rotate()
```

### Description One

Most of the loans in the dataset are currently active (56,000). Over 38,000,
was paid off. Around 16 thousand were charged off or defaulted. Understanding
what reasons stay behind the defaulted and charged off loans is very important
from the loan safety perspective. Both investors and loan offering companies
are highly interested in understanding what factors impact the borrower's 
ability to pay off the debt.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(StatedMonthlyIncome, MonthlyLoanPayment), data=df) +
  geom_jitter(aes(color=LoanStatus),alpha=1) +
  scale_x_continuous(limits = c(0.0, quantile(df$StatedMonthlyIncome, 0.99)),
                     breaks = seq(0, 100000, 1000),
                     labels=dollar_format(prefix="$")) +
  scale_y_continuous(limits = c(0.0, quantile(df$MonthlyLoanPayment, 0.99)),
                     breaks = seq(0, 1500, 100),
                     labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = "Stated Monthly Income, Monthly Loan Payment and 
Loan Status relation") +
  xlab("Stated Monthly Income") +
  ylab("Monthly Loan Payment") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(angle = 90, hjust = 1, size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

### Description Two

In this plot, we can look closer at the stated monthly income, monthly loan 
payment and loan status. It is interesting to see a lot of charged-off and 
defaulted loans in the lower left corner of the plot - small monthly payment
but a small income. It seems that borrowers with higher income can pay off 
the loan even if the monthly payment is larger, while borrowers with the 
low-income struggle to pay off the loan even if the monthly payment is low.


### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(ProsperRating..Alpha., LoanOriginalAmount), data=df) +
  geom_boxplot(aes(fill=LoanStatus),alpha=1) +
  scale_y_continuous(breaks = seq(0, 50000, 2500),
                     labels=dollar_format(prefix="$")) +
  guides(fill = guide_legend(override.aes= list(alpha = 1))) +
  scale_color_manual(values=c("#539BBA", 
                              "#408B5C", 
                              "#C0BE59", 
                              "#C05E59", 
                              "#664081", 
                              "#9B4875", 
                              "#686977")) +
  labs(title = 
         "Prosper Rating, Loan Original Amount and Loan Status relation") +
  ylab("Loan Original Amount") +
  xlab("Prosper Rating") +
  theme(plot.title = element_text(size = 18),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=10),
        panel.grid.major.x = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5),
        panel.grid.major.y = element_line(color="#d3d3d3", 
                                          linetype = "dashed", 
                                          size=0.5))
```

### Description Three

In this plot, we can look closer at the prosper rating, loan original amount
and loan status. We can see that most of the completed loans were loans with
a small amount.  Most of the loans with a high amount are still active. 
In this plot, we can also clearly see a relation between prosper score 
and the amount of loan. Borrowers with a low score are granted smaller loans,
while those with good rating bigger loans. High Rish and E category borrowers
are almost never granted a loan above the \$10,000, while we can see that some
of the AA, A and B category borrowers have loans above the \$25,000.

------

# Reflection

I have tried to understand if employment status, loan amount, monthly loan
payment and monthly income have a significant impact on the borrower's ability
to pay off the loan.  The  main challenge I have faced with this dataset 
is that for lots of loans we did not have all information e.g. employment 
status, listening category was missing. This has significantly limited 
the investigation and ability do drive conclusions out of data. Gathering
more data about the borrowers could help us understand better the relations 
between loan status and borrower's profile.